---
title: "my-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(EGS2)

```

# **Introduction**

Undergoing an emergency general surgery (EGS) procedure increases odds of death by 40% and major complications by 30% compared to performing the same procedure on an elective basis. While the emergency nature will likely always increase the risks of the operation, nationally organized quality improvement efforts in emergency general surgery are lacking compared to other surgical areas such as trauma, elective colorectal, and orthopedic surgery, despite the fact that these surgeries carry some of the highest risk.  The Acute Care Surgery Group at Indiana University has been implementing new and innovative protocols to address this gap and improve outcomes. One portion of this project is improving time to OR for the three specific indications associated with the highest case fatality rates: bowel ischemia, intestinal perforation, and necrotizing soft tissue infection.

To help accomplish this, an EGS Alert system was created for the Emergency Department (ED).  If the patient is a transfer from an outside hospital, an emergency page is pushed to the surgery team on arrival.  The attending surgeon must see the patient within 15 minutes to make an operative decision.  If the patient is a primary presentation to the ED, they are paged out on arrival if they are unstable with suspicion of intraabdominal cause or, if stable, after a diagnosis of one of these conditions is made by CT scan.  Additionally, a quality target of two hours from diagnosis to OR was chosen regardless of whether an EGS alert was activated or not. The two hour goal is defined as two hours from arrival to OR for transfer patients (whose diagnosis is already known) or from CT scan to OR for patients who present primarily to the ED (whose diagnosis is unknown on arrival).

Outcome and process improvement data is reviewed monthly at the ACS Operations Meeting. However, the data review and creation of the meeting presentation can be very time-consuming and detract from other important efforts.  Currently, the data is pulled from the Emergency General Surgery (EGS) registry into a bulk Excel file.  This is then manually examined, calculations are made, and graphical representations are created in Excel and then copied to PowerPoint.  This can be quite time consuming.

The aim of this package is to input the raw data file, tidy the data, calculate the appropriate time to OR, and create the graphics for the presentation, saving a substantial amount of time over the course of a year.

To accomplish this, the user imports a data file from the registry. The column names must match the default output of the EGS registry and be labeled as "Admission Date/Time", "Surgical Procedure?", "First CT Date/Time", "First OR Date/Time", "Primary Surgical Dx Code", and "Transfer in?". Since the package is targeted to users of this specific database, this should not be an issue unless the user has manually changed the name.

The data frame is then simply inputted into the EGS() function which automatically edits the data to include only the three types of cases noted above, and then calculates the time to OR appropriately based on the transfer/nontransfer status.  The time to OR will then be plotted for each of the three surgical indications. The average will be presented for each month of the year to date.  A shiny gadget allows the user to graphically exclude outliers. 

An included R markdown file titled "EGSOps.Rmd" in the vignettes folder allows the code to be directly used to make powerpoint slides.

In summary, there are three challenges that the software addresses:

1. Reducing workload of data staff by generating the operations slides for analysis with a simple command rather than hours of manual data manipulation.

2. Allowing rapid, simple exclusion of outliers by the user through a shiny gadget rather than manually editing the excel spreadsheet

3. Ability to do more quality graphical representations than what has previously been made by the data team without adding time to the process.

# **Examples**
```{r, echo = TRUE, message=FALSE, warning=FALSE}
#Packages required to run the EGS package
library(readxl)
library(dbplyr)
library(ggplot2)
library(tidyverse)
library(officedown)
library(regexSelect)
library(lubridate)
library(knitr)
library(roxygen2)
library(shiny)
library(miniUI)
```


We will read in this example file using the readxl package.

```{r}
library(readxl)
X2020_EGS_Quality_Data <- read_excel("~/Documents/MS Biostatistics/B581 Biostats Computing/Final Project/2020 EGS Quality Data.xlsx")

```

If the package has been loaded, we can now utilize the function EGS() in the console
One can simply enter in the console "EGS(X2020_EGS_Quality_Data)"

Individual plots can then be called and exported as the user would like.

However, to show how this function works, we are going to go through the code within the function step-by-step.

First, the data is filtered to only include the columns that are needed. The nonsurgical cases are removed, and the dates are converted using the lubridate package.  The actual function "EGS" allows the user to input a data file with any title into the function. For this vignette example, we have manually called the example file.

```{r, warning=FALSE}
DS1 <-X2020_EGS_Quality_Data %>% select("Admission Date/Time", "Surgical Procedure?", "First CT Date/Time", "First OR Date/Time", "Primary Surgical Dx Code", "Transfer in?")
DS1NS <- DS1 %>% filter(`Surgical Procedure?`=="YES")
DS1NST <- DS1NS %>% mutate(`First CT Date/Time` = mdy_hm(`First CT Date/Time`)) %>% mutate(`First OR Date/Time` = mdy_hm(`First OR Date/Time`))
DS1NSTNA <- DS1NST %>% drop_na(`First OR Date/Time`)
```

Next, two new variables are calculated, AOR and CTOR, which are the admission to OR times and the CT to OR times. Lubridate is used, and the result in hours is rounded to two decimal places. Subsequently, ifelse terminology is used to select the admission to OR or the CT to OR time based on the transfer status of the patient and placed in the new column TTOR. Additionally, a column called Month is added in order to facilitate plotting later.

```{r}
DS1NSTNA$AOR <- round(difftime(DS1NSTNA$`First OR Date/Time`, DS1NSTNA$`Admission Date/Time`, units="hours"), 2)
DS1NSTNA$CTOR <- round(difftime(DS1NSTNA$`First OR Date/Time`, DS1NSTNA$`First CT Date/Time`, units="hours"), 2)
DS1NSTNA <- DS1NSTNA %>% drop_na(AOR)
DS1NSTNA <- DS1NSTNA %>% drop_na(CTOR)
DS1NSTNA$TTOR <- ifelse(DS1NSTNA[,6] == "YES", DS1NSTNA$AOR, DS1NSTNA$CTOR)
DS1NSTNA$Month <- month(DS1NSTNA$`First OR Date/Time`, label = TRUE, abbr = FALSE)
```

Now, in any dataset, there are always outliers.  The shiny gadget titled "pick_points" is used to allow the user to visually exlude outliers.  Unfortunately, R tends to freeze in package creation when the shiny code is included. The example is included below but is not active in the vignette to prevent freezing. If you utilize the actual function or the r markdown powerpoint file included in the package, you can manually pick points with the shiny gadget.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#pick_points <- function(data, x, y) {
#  ui <- miniPage(
#    gadgetTitleBar(paste("Select points")),
#    miniContentPanel(padding = 0,
#      plotOutput("plot1", height = "100%", brush = "brush")
#    ),
#    miniButtonBlock(
#      actionButton("add", "", icon = icon("thumbs-up")),
#      actionButton("sub", "", icon = icon("thumbs-down")),
#      actionButton("none", "" , icon = icon("ban")),
#      actionButton("all", "", icon = icon("refresh"))
#    )
#  )

#  server <- function(input, output) {
    # For storing selected points
#    vals <- reactiveValues(keep = rep(TRUE, nrow(data)))

#    output$plot1 <- renderPlot({
      # Plot the kept and excluded points as two separate data sets
#      keep    <- data[ vals$keep, , drop = FALSE]
#      exclude <- data[!vals$keep, , drop = FALSE]

#      ggplot(keep, aes_(x, y)) +
#        geom_point(data = exclude, color = "grey80") +
#        geom_point()
#    })

    # Update selected points
#    selected <- reactive({
#      brushedPoints(data, input$brush, allRows = TRUE)$selected_
#    })
#    observeEvent(input$add,  vals$keep <- vals$keep | selected())
#    observeEvent(input$sub,  vals$keep <- vals$keep & !selected())
#    observeEvent(input$all,  vals$keep <- rep(TRUE, nrow(data)))
#    observeEvent(input$none, vals$keep <- rep(FALSE, nrow(data)))

#    observeEvent(input$done, {
#      stopApp(vals$keep)
#    })
#    observeEvent(input$cancel, {
#      stopApp(NULL)
#    })

#  }

#  runGadget(ui, server)
#}
#DS1NSTNA$Selected <- pick_points(DS1NSTNA, ~`First OR Date/Time`, ~TTOR)
```

The dataset is then edited to filter out the points as selected by the user. We have converted the true code to text and added a manual filter due to the limitation of the vignette to run shiny during package creation in order to show what the plots would look like after the user filters the outliers.

A column named Cat is created to facilitate plotting based on diagnosis code category. 

```{r}
#DS1NSTNAS <- DS1NSTNA %>% filter(Selected ==TRUE)
DS1NSTNAS <- DS1NSTNA %>% filter(TTOR <= 10)
NSTISet <- DS1NSTNAS %>% filter(`Primary Surgical Dx Code`  %in% c("M72.6", "N49.3"))
NSTISet$Cat <- "NSTI"
IschSet <- DS1NSTNAS %>% filter(str_detect(`Primary Surgical Dx Code`, "^K55"))
IschSet$Cat <- "ISCH"
PerfSet <- DS1NSTNAS %>% filter(str_detect(`Primary Surgical Dx Code`, "^K63|K25|K26|K57"))
PerfSet$Cat <- "PERF"
FinalSet <- rbind(NSTISet, IschSet, PerfSet)

```

Now, the plots are created using ggplot, which is our desired final output.

```{r}
boxp <- ggplot(FinalSet, aes(Cat, as.numeric(TTOR))) + geom_boxplot(fill="steelblue", outlier.shape = NA) + labs(title = "Time to OR by Diagnosis", x= "Diagnosis", y= "Time to OR in Hours") + theme_light() + theme(plot.title = element_text(hjust = 0.5)) + geom_jitter(width=0.05, color = "red") 
boxp
```

```{r}
avtimes <- as.numeric(c(round(mean(NSTISet$TTOR), 2), round(mean(IschSet$TTOR), 2), round(mean(PerfSet$TTOR),2)))
titles <- c("NSTI", "Ischemia", "Perforation")
avbydx <- cbind.data.frame(titles, avtimes)

Plotavbydx <- ggplot(avbydx, aes(x = titles,y = avtimes)) +geom_col(fill="steelblue") + labs(title = "YTD Average Time to OR", x= "Diagnosis", y= "Time to OR in Hours") + theme(plot.title = element_text(hjust = 0.5)) + geom_text(aes(label=avtimes), vjust=-0.3, size=3.5)+ theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
Plotavbydx

```

```{r}
avbymonth <- aggregate(x=FinalSet$TTOR, by =  list(FinalSet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonth
plotavbymonth <- ggplot(avbymonth, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="red", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR by Month", y= "Time to OR in Hours") +  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=0, nudge_y = .5)
plotavbymonth
```

```{r}
avbymonthNSTI <- aggregate(x=NSTISet$TTOR, by = list(NSTISet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonthNSTI
plotavbymonthNSTI <- ggplot(avbymonthNSTI, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="green", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR for NSTI", y= "Time to OR in Hours") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=-0.2)
plotavbymonthNSTI

```

```{r}
avbymonthIsch <- aggregate(x=IschSet$TTOR, by = list(IschSet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonthIsch
plotavbymonthIsch <- ggplot(avbymonthIsch, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="green", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR for Ischemia", y= "Time to OR in Hours") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=-0.2)
plotavbymonthIsch

boxp <- ggplot(FinalSet, aes(Cat, as.numeric(TTOR))) + geom_boxplot(fill="steelblue", outlier.shape = NA) + labs(title = "Time to OR by Diagnosis", x= "Diagnosis", y= "Time to OR in Hours") + theme_light() + theme(plot.title = element_text(hjust = 0.5)) + geom_jitter(width=0.05, color = "red") 
boxp
```

```{r}
avbymonthPerf <- aggregate(x=PerfSet$TTOR, by = list(PerfSet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonthPerf
plotavbymonthPerf <- ggplot(avbymonthPerf, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="green", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR for Perforations", y= "Time to OR in Hours") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=-0.2)
plotavbymonthPerf
```

As a second example, we will show how the plots would appear without outlier removal from this data sample. This will exhibit the benefit of the built-in shiny gadget.  We have not repeated the code, which is unchanged from above except for not filtering the outliers.

```{r, echo = FALSE}
NSTISet <- DS1NSTNA %>% filter(`Primary Surgical Dx Code`  %in% c("M72.6", "N49.3"))
NSTISet$Cat <- "NSTI"
IschSet <- DS1NSTNA %>% filter(str_detect(`Primary Surgical Dx Code`, "^K55"))
IschSet$Cat <- "ISCH"
PerfSet <- DS1NSTNA %>% filter(str_detect(`Primary Surgical Dx Code`, "^K63|K25|K26|K57"))
PerfSet$Cat <- "PERF"
FinalSet <- rbind(NSTISet, IschSet, PerfSet)

boxp <- ggplot(FinalSet, aes(Cat, as.numeric(TTOR))) + geom_boxplot(fill="steelblue", outlier.shape = NA) + labs(title = "Time to OR by Diagnosis", x= "Diagnosis", y= "Time to OR in Hours") + theme_light() + theme(plot.title = element_text(hjust = 0.5)) + geom_jitter(width=0.05, color = "red") 
boxp

avtimes <- as.numeric(c(round(mean(NSTISet$TTOR), 2), round(mean(IschSet$TTOR), 2), round(mean(PerfSet$TTOR),2)))
titles <- c("NSTI", "Ischemia", "Perforation")
avbydx <- cbind.data.frame(titles, avtimes)

Plotavbydx <- ggplot(avbydx, aes(x = titles,y = avtimes)) +geom_col(fill="steelblue") + labs(title = "YTD Average Time to OR", x= "Diagnosis", y= "Time to OR in Hours") + theme(plot.title = element_text(hjust = 0.5)) + geom_text(aes(label=avtimes), vjust=-0.3, size=3.5)+ theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
Plotavbydx

avbymonth <- aggregate(x=FinalSet$TTOR, by =  list(FinalSet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonth
plotavbymonth <- ggplot(avbymonth, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="red", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR by Month", y= "Time to OR in Hours") +  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=0, nudge_y = .5)
plotavbymonth

avbymonthNSTI <- aggregate(x=NSTISet$TTOR, by = list(NSTISet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonthNSTI
plotavbymonthNSTI <- ggplot(avbymonthNSTI, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="green", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR for NSTI", y= "Time to OR in Hours") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=-0.2)
plotavbymonthNSTI

avbymonthIsch <- aggregate(x=IschSet$TTOR, by = list(IschSet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonthIsch
plotavbymonthIsch <- ggplot(avbymonthIsch, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="green", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR for Ischemia", y= "Time to OR in Hours") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=-0.2)
plotavbymonthIsch

boxp <- ggplot(FinalSet, aes(Cat, as.numeric(TTOR))) + geom_boxplot(fill="steelblue", outlier.shape = NA) + labs(title = "Time to OR by Diagnosis", x= "Diagnosis", y= "Time to OR in Hours") + theme_light() + theme(plot.title = element_text(hjust = 0.5)) + geom_jitter(width=0.05, color = "red") 
boxp

avbymonthPerf <- aggregate(x=PerfSet$TTOR, by = list(PerfSet$Month), FUN = mean) %>% rename(Month=Group.1, Average = `Transfer in?`)
#avbymonthPerf
plotavbymonthPerf <- ggplot(avbymonthPerf, aes(x = Month, y = Average, group = 1, label = round(Average, digits = 2))) + geom_line(color="green", size=1.2) + geom_point(color = "blue", size = 2) + labs(title = "Average Time to OR for Perforations", y= "Time to OR in Hours") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) + geom_text(nudge_x=-0.2)
plotavbymonthPerf


```

For the vast majority of large outliers, the diagnosis was typically not made at the time of admission or visible on the initial CT scan. By skewing the analysis with outliers, it is difficult to look at the remainder of the data, as you can see by comparing the plots before and after outlier removal. By allowing the user to rapidly remove the outliers utilizing shiny, the plots generated are much more helpful for understanding the data and much more rapidly created that manually editing excel files and generating plots with Office.




# **Future Work**

While this function successfully generates the plots looking at our metrics for time to OR for our three targeted diagnoses, there are other metrics/slides that we currently use each month in our operations meeting, such as overall volume, time to discharge, etc.  Future work will be to make the entire Operations powerpoint in r.  Additionally, it will be beneficial to allow selection and analysis for each subgroup utilizing Shiny's ability to brush points. In its current version, we remove outliers for all case types at one time.  We also plan to add additional functionality for presenting the outliers in table format so that they are easily identified for review. The current examples do not utilize any patient identifiers due to HIPAA. Once this function is being used internally, we would be able to include identifiers to easily flag patients that need review.
